\documentclass[12pt]{article}

% --- Formatting per candidacy requirements -------------------------------
% Proposition limit: 15 double-spaced pages, excluding front matter
% and references.
% This is the OUT-OF-FIELD proposition.
% -------------------------------------------------------------------------


\usepackage[letterpaper,margin=1in]{geometry}
\usepackage{setspace}
\doublespacing

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}

\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{physics}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{xcolor}
\usepackage[hidelinks]{hyperref}
\usepackage[capitalize,noabbrev]{cleveref}

\usepackage[
  backend=biber,
  style=numeric-comp,
  sorting=none,
  giveninits=true,
  maxbibnames=10,
  url=false
]{biblatex}
\addbibresource{Quantum Crypto/Quantum Crypto.bib}

\graphicspath{{figures/}}

\title{Proposition 2 (Out-of-Field) \\[0.25em]
       \large Public-Key Quantum Money}
\author{Xiao-Yu Ouyang}
\date{\today}

\begin{document}

% --- Front matter (excluded from page count) -----------------------------
\begin{titlepage}
  \centering
  \vspace*{2cm}
  {\Large \bfseries Candidacy Proposition 2\par}
  \vspace{0.5cm}
  {\large \textit{Out-of-field}\par}
  \vspace{1.5cm}
  {\LARGE Public-Key Quantum Money\par}
  \vspace{2cm}
  {\large Xiao-Yu Ouyang \par}
  \vspace{0.5cm}
  {\normalsize \texttt{xouyang@caltech.edu} \par}
  \vspace{1.5cm}
  {\normalsize Division of Chemistry and Chemical Engineering \\
               California Institute of Technology \par}
  \vfill
  {\normalsize Submitted \today \par}
\end{titlepage}

\begin{singlespace}
\tableofcontents
\end{singlespace}
\newpage

\begin{abstract}
\noindent

\end{abstract}

% --- Body (counts toward the 15-page limit) ------------------------------
\setcounter{page}{1}

\section{Introduction}
Quantum money originated with Wiesner's proposal of conjugate coding, where classical information is encoded into quantum states chosen from incompatible bases \cite{wiesnerConjugateCoding1983}.
In the simplest form, the bank chooses secret strings $x,\theta \in \{0,1\}^n$ and prepares
\[
|\$_s\rangle = |s\rangle \otimes \bigotimes_{i=1}^n H^{\theta_i}|x_i\rangle,
\]
where $s$ is a classical serial number and $H^{\theta_i}$ means that the $i$th qubit is prepared either in the computational basis $\{|0\rangle,|1\rangle\}$ or in the Hadamard basis $\{|+\rangle,|-\rangle\}$.
These are conjugate, or mutually unbiased, bases: measuring a qubit in the wrong basis gives a random outcome and disturbs the state.
The bank can verify the note because it stores the secret classical description $(x,\theta)$ associated with the serial number $s$, while a counterfeiter who does not know this description cannot learn the note well enough to reproduce it without disturbing it.
This was the first example of a cryptographic task whose security is genuinely quantum, since a classical string can always be copied, but an unknown quantum state cannot be cloned.
Wiesner's scheme, however, is private-key quantum money.
This means that verification requires returning the banknote to the bank, or at least interacting with an entity that knows the secret verification data.

A natural next goal is to make quantum money behave more like cash.
In a public-key quantum money scheme, anyone should be able to verify a banknote using a classical public key and a public description of a quantum verification procedure, while only the legitimate issuer should be able to mint valid notes efficiently.
Early work of Bennett, Brassard, Breidbart, and Wiesner already explored ``unforgeable subway tokens'': quantum tokens that a turnstile can verify, but that a user should not be able to copy and reuse \cite{bennettQuantumCryptographyUnforgeable1982}.
Aaronson later gave a complexity-theoretic formulation of public quantum money and proved a complexity-theoretic no-cloning theorem, giving oracle evidence that public verification need not make cloning easy \cite{aaronsonQuantumCopyProtectionQuantum2009}.
In general, the honest mint has a secret key or trapdoor and can prepare valid banknotes efficiently, while the security requirement is that a public adversary, given the public key, the public verifier, and some valid banknotes, cannot efficiently prepare or clone additional valid banknotes.

Many candidate public-key quantum money schemes can be understood by asking what the banknote state is, how public verification works, and what computational assumption prevents public state preparation.
Aaronson's early candidate used random stabilizer states and public tests, but this concrete proposal was later broken \cite{aaronsonQuantumCopyProtectionQuantum2009,lutomirskiBreakingMakingQuantum2009}.
Farhi, Gosset, Hassidim, Lutomirski, and Shor proposed money from knots, where a banknote is a superposition over equivalent knot diagrams and verification uses knot invariants \cite{farhiQuantumMoneyKnots2012}.
Its security is heuristic and topological: the intended hardness is that, given public invariants such as the Alexander polynomial, it is still hard to prepare the required superposition over equivalent diagrams.
Aaronson and Christiano proposed a hidden-subspace scheme whose verifier performs tests in complementary bases \cite{aaronsonChristianoQuantumMoneyHidden2012}.
The black-box version has strong security evidence, but making the hidden subspace publicly testable requires concrete algebraic encodings; later cryptanalysis showed that some proposed concrete hard problems for such schemes are easier than expected \cite{bilykCryptanalysisThreeQuantum2022}.
Zhandry introduced quantum lightning, a stronger ``collision-free'' primitive in which no efficient procedure should generate two valid bolts with the same serial number \cite{zhandryQuantumLightningNever2017a}.
Lightning is closely related to public-key quantum money, but it changes the minting model: generation may be public, while security forbids serial-number collisions rather than merely counterfeiting notes issued by a bank.
The mechanism is cryptographic rather than statistical: the serial number is tied to the quantum state by the verification relation, and producing two valid states with the same serial number should solve the underlying hard problem.

The search for assumptions underlying public verification has produced both constructions and barriers.
Liu, Montgomery, and Zhandry showed that a natural lattice-based route to public quantum money cannot be secure, including an attack on a scheme of Khesin, Lu, and Shor \cite{liuAnotherBreakingMaking2023}.
They also proposed invariant money, a framework abstracting the structure of knot-based money and relying on new knowledge-type assumptions; possible instantiations include group actions, rerandomizable functional encryption, isogenies, and knots.
Kane, Sharif, and Silverberg proposed a number-theoretic approach in which bills are joint eigenstates of commuting operators arising from quaternion algebras \cite{kaneQuantumMoneyQuaternion2022}.
Other works show barriers to obtaining public-key quantum money from some standard-looking tools, including collision-resistant hash functions and black-box uses of evasive obfuscation \cite{ananthImplausibilityPublicKey2023,zhandryQuantumMoneyEvasive2025}.
There is also a semi-quantum direction, initiated by Radian and Sattath, that asks how much quantumness is needed from the bank; Shmueli constructed public-key quantum money with a classical bank from quantum-secure indistinguishability obfuscation and sub-exponential LWE \cite{radianSemiQuantumMoney2019,shmueliPublicKeyQuantum2022}.
Zhandry constructed public-key quantum money and quantum lightning from abelian group actions \cite{zhandryQuantumMoneyAbelian2025}; 
Doliskani related the security of this construction in a generic group-action model to the group-action discrete logarithm problem \cite{doliskaniPublicKeyQuantum2025}; 
and Bostanci, Nehoran, and Zhandry extended the picture through a general quantum duality for group representations and a pre-action security assumption \cite{bostanciGeneralQuantumDuality2025}.

These works suggest a common verification-preparation tradeoff.
Existing public-key quantum money schemes have efficient public verification by definition, but the verifier is usually described as a public quantum circuit or an algebraic test.
This proposal asks whether the same tradeoff can be expressed in the language of Hamiltonian complexity.
Instead of viewing verification only as a public verification circuit, we ask whether the valid banknotes can be realized as low-energy states, or ground states, of a publicly specified Hamiltonian $H$.
In such a formulation, the public verifier is an energy test for $H$, while the security question becomes a ground-state preparation question: can an adversary given $H$, $\mathsf{pk}$, and $k$ valid ground states efficiently prepare more valid ground states?

The Hamiltonian viewpoint is natural because circuit verification can often be converted into local Hamiltonian constraints, and geometrically local Hamiltonians are the standard bridge between quantum computation and many-body physics \cite{gottesmanQuantumClassicalComplexity2010}.
Recent work on public-key pseudoentanglement and ground-state learning hardness also shows that cryptographic assumptions can control what can be learned or prepared from a public Hamiltonian description \cite{boulandPublickeyPseudoentanglementHardness2023,boulandHardnessLearningGround2024}.
The main question is simple: can public-key quantum money be realized as the ground space of a public local Hamiltonian?
The desired construction should keep the honest mint efficient, while making ground-state preparation hard for anyone with only public information.

\subsection{Definition}
A public-key quantum money scheme consists of an efficient minting algorithm and an efficient public verification algorithm:
\begin{itemize}
    \item $\mathrm{Gen}(\mathsf{sk}) \to (\mathsf{pk}, |\$\rangle)$: Given a classical secret key $\mathsf{sk}$, the minting algorithm outputs a classical public key $\mathsf{pk}$ and a quantum banknote state $|\$\rangle$.
    \item $\mathrm{Ver}(|\psi\rangle,\mathsf{pk}) \to \top / \bot$: Given a candidate quantum state $|\psi\rangle$ and the classical public key $\mathsf{pk}$, the public verification algorithm accepts or rejects the state.
\end{itemize}

\subsection{Security (Unclonability)}
The basic security requirement is that one valid banknote cannot be turned into two valid banknotes.
Let $\mathcal{A}$ be a quantum polynomial-time adversary that receives the public key $\mathsf{pk}$ and one valid banknote $|\$\rangle$.
The adversary applies an arbitrary quantum attack and outputs a joint state on two registers, which are intended to be two candidate banknotes:
\[
\rho_{A_1A_2} \leftarrow \mathcal{A}(\mathsf{pk},|\$\rangle).
\]
Here $\rho_{A_1A_2}$ may be mixed or entangled across the two registers.
The individual candidate banknotes are the reduced states obtained by tracing out the other register:
\[
\rho_{A_1}=\operatorname{Tr}_{A_2}(\rho_{A_1A_2}), \qquad
\rho_{A_2}=\operatorname{Tr}_{A_1}(\rho_{A_1A_2}).
\]
Unclonability says that the probability that both candidate banknotes pass public verification is negligible:
\[
\Pr\!\left[
\mathrm{Ver}(\rho_{A_1},\mathsf{pk})=\top
\ \land\
\mathrm{Ver}(\rho_{A_2},\mathsf{pk})=\top
\right]
\le \operatorname{negl}(\lambda).
\]
Equivalently, no efficient adversary should be able to increase the number of valid banknotes.
The fully general definition allows an adversary to receive $k$ valid banknotes and output more than $k$ registers, and requires that the adversary's total expected number of accepting registers increases by at most a negligible amount.

\subsection{Constructions}
There are many proposed candidate constructions of public-key quantum money.
For this proposal, two representative families are especially relevant: constructions based on post-quantum indistinguishability obfuscation, and constructions based on cryptographic group actions.
Both families try to solve the same basic problem: publish enough information to verify a state, while hiding enough structure so that the public information does not become a state-preparation algorithm.

\paragraph{Post-quantum indistinguishability obfuscation.}
Indistinguishability obfuscation (iO) is a way to publish a circuit while hiding how the circuit works.
Informally, if two circuits compute the same function, then their obfuscations should be computationally indistinguishable:
\[
C_0(x)=C_1(x)\ \text{for all }x
\quad \Longrightarrow \quad
\mathsf{iO}(C_0)\approx_{\mathrm{q}}\mathsf{iO}(C_1),
\]
where $\approx_{\mathrm{q}}$ means indistinguishability for efficient quantum adversaries.
In public-key quantum money, iO can be used to publish a verification circuit that checks membership in a hidden valid-state space without revealing the trapdoor used by the mint.
Schematically, the public key contains an obfuscated verifier,
\[
\mathsf{pk}=\mathsf{iO}(V_{\mathsf{sk}}),
\qquad
\mathrm{Ver}(\rho,\mathsf{pk})=V_{\mathsf{pk}}(\rho),
\]
while the secret key $\mathsf{sk}$ is still needed to efficiently prepare valid banknotes.
This route appears in constructions and proposals based on quantum-secure iO, often together with additional assumptions such as sub-exponential LWE \cite{zhandryQuantumLightningNever2017a,shmueliPublicKeyQuantum2022}.

\paragraph{Group actions.}
The group-action approach gives a more direct bridge to projectors and Hamiltonians.
Let a finite group $G$ act on a finite set $X$, written $g \star x$.
This action defines a unitary representation on the Hilbert space spanned by $\{\ket{x}:x\in X\}$:
\[
R(g)=\sum_{x\in X}\ket{g\star x}\bra{x}.
\]
For an irreducible representation $\lambda$ of $G$, with dimension $d_\lambda$ and character $\chi_\lambda$, the corresponding Fourier-isotypic projector is
\[
P_\lambda=\frac{d_\lambda}{|G|}
\sum_{g\in G}\chi_\lambda(g)^*R(g).
\]
In the group-action quantum money picture, the serial number is associated with $\lambda$, and a valid banknote is a state in the image of $P_\lambda$.
Public verification can therefore be viewed as checking whether the state has large overlap with this projector:
\[
\mathrm{Ver}(\rho,\lambda)=\top
\quad \text{if} \quad
\operatorname{Tr}(P_\lambda \rho)\approx 1.
\]
This immediately gives a Hamiltonian whose ground space is the valid money subspace,
\[
H_\lambda=I-P_\lambda.
\]
The security of this route is based on assumptions about the hardness of manipulating the group action, such as generic group-action hardness or pre-action security \cite{zhandryQuantumMoneyAbelian2025,doliskaniPublicKeyQuantum2025,bostanciGeneralQuantumDuality2025}.

\paragraph{Circuit-to-Hamiltonian Transformation:}
Both representative routes provide public verification procedures, either as circuits or as projectors.
The Feynman-Kitaev circuit-to-Hamiltonian construction can transform a verification circuit $V$ into a local Hamiltonian whose ground states are history states of the computation.
However, this does not automatically imply that the banknote itself is the ground state.
For group actions, the projector formula above gives the non-local Hamiltonian $H_\lambda=I-P_\lambda$ directly; the main challenge is to make an analogous Hamiltonian local or geometrically local while preserving mintability and unclonability.

\paragraph{Open Question:}
For the group-action construction, can the verification projector, or an equivalent verifier, be transformed into a \textbf{1D/2D geometrically local Hamiltonian} whose ground states remain valid public-key quantum money states \cite{gottesmanQuantumClassicalComplexity2010}?

---

\section{Complexity Classes: QMA vs. Clonable QMA}

\subsection{QMA (Quantum Merlin-Arthur)}
A language $L \in \text{QMA}$ iff there exists a verification procedure $V$ such that for an input $x$:
\begin{itemize}
    \item \textbf{Completeness:} If $x \in L$, $\exists \ket{\psi}$ such that $\Pr[V(x, \ket{\psi}) \to \top] \ge 2/3$.
    \item \textbf{Soundness:} If $x \notin L$, $\forall \ket{\psi}$, $\Pr[V(x, \ket{\psi}) \to \top] \le 1/3$.
\end{itemize}

\subsection{Clonable QMA}
This class deals with problems where the witness is explicitly "clonable" or the verifier receives multiple copies.
\begin{itemize}
    \item Distinguishing between cases where a witness $\ket{\psi}$ can verify with high probability vs. cases where no witness works.
    \item In Clonable QMA, we consider if the witness $\ket{\psi}$ allows for $\ket{\psi} \to \ket{\psi}\ket{\psi}$.
\end{itemize}

\subsection{QMA-Complete Problem: Local Hamiltonian}
Given a $k$-local Hamiltonian $H$, decide the ground state energy $E_0$:
\begin{itemize}
    \item YES case: $E_0 \le a$
    \item NO case: $E_0 \ge b$
    \item Promise gap: $b - a = \frac{1}{\text{poly}(n)}$
\end{itemize}

\paragraph{Question:} Can we find a local Hamiltonian whose ground state is \textbf{unclonable}?

\subsection{Separation Result}
There is a separation between \textbf{QMA} and \textbf{Clonable QMA} (citing Bostanci, Haferkamp, Nirkhe, Zhandry). This involves separating QMA and QCMA with a classical oracle.

\subsubsection{Problem: Spectral Correlation}
Given sets $S, T \subseteq \{0,1\}^n$, define projections:
\[
\Pi_S = \sum_{x \in S} \ket{x}\bra{x}, \quad \Pi_T = \sum_{x \in T} \ket{x}\bra{x}
\]
Decide if there exists $\ket{\psi}$ such that:
\[
\| \Pi_S H^{\otimes n} \Pi_T \ket{\psi} \|^2 \ge \frac{2}{3} \quad \text{(YES case)}
\]
Or if for all $\ket{\psi}$, the value is $\le \frac{1}{3}$ (NO case).

---

\section{Group Actions and Oracles}

\subsection{Search for Oracle Separation}
Can we find a classical oracle relative to which there exists a Hamiltonian $H$ such that:
\begin{itemize}
    \item It is easy to decide its ground state energy in QMA.
    \item It is hard to find a clonable witness (or the ground state is hard to clone).
\end{itemize}

\subsection{Pre-action Security of Group Action}
Let $G$ be a group acting on set $X$, denoted $g \star x$.
\begin{itemize}
    \item \textbf{Pre-action:} For an element $h \in G$, considering actions like $g \star h \star x$ vs $g \star x$.
    \item \textbf{Security Definition:} It is hard to distinguish the case where a random action has been applied to a register from the case where a random \textbf{pre-action} and action have been applied.
    \item Formally, distinguishing:
    \[ h_1 g_1 x \quad \text{vs} \quad h_1 h_2 \dots \quad (\text{where } h_i, g_i \leftarrow G) \]
\end{itemize}

\paragraph{Instantiations \& Applications:}
\begin{itemize}
    \item Can be instantiated with \textbf{McEliece} (code-based crypto).
    \item Leads to \textbf{One-Shot Signatures} and \textbf{Quantum Money}.
    \item Connections to Classical Oracles, iO, and sub-exponential LWE.
\end{itemize}
% --- References (excluded from page count) -------------------------------
\begin{singlespace}
\printbibliography
\end{singlespace}
\end{document}
